% ============================================================================
% ICSE Paper — Approach Section
% LLM-Driven Trace Link Recovery between Architecture Documents and Models
% ============================================================================

\section{Approach}\label{sec:approach}

\subsection{Overview}\label{sec:overview}

We propose an LLM-driven pipeline for recovering trace links between
Software Architecture Documents (SAD) and Software Architecture Models (SAM).
Given a natural-language architecture document~$D$ consisting of sentences
$\{s_1, \ldots, s_n\}$ and a component model~$M$ containing components
$\{c_1, \ldots, c_m\}$, our approach produces a set of trace links
$L \subseteq S \times C$, where each link $(s_i, c_j)$ asserts that
sentence~$s_i$ discusses component~$c_j$.

The pipeline is organized as a \emph{fan-out / fan-in} staged DAG
(Figure~\ref{fig:pipeline}) with three layers:

\begin{enumerate}
    \item \textbf{Knowledge acquisition} (Layer~1): Independent analyses
      extract structured knowledge from the raw inputs---model structure,
      document vocabulary, and linguistic patterns.
      Model analysis, document analysis, and explicit reference
      extraction share no data dependencies; pattern learning and
      multiword enrichment depend on model analysis results and run
      once the initial analyses complete.
    \item \textbf{Link recovery} (Layer~2): Four complementary strategies
      produce candidate links, each targeting a distinct reference type:
      \emph{explicit} references (verbatim names), \emph{contextual}
      references (aliases, functional descriptions), \emph{anaphoric}
      references (pronouns), and \emph{abbreviated} references (partial
      names).
      The strategies are largely independent, and their outputs converge
      only at the merge point.
    \item \textbf{Link consolidation} (Layer~3): A single fan-in stage
      merges all candidate links, applies convention-aware boundary
      filtering to \emph{all} links regardless of provenance, and removes
      residual false positives via a deterministic evidence filter.
\end{enumerate}

\noindent This architecture offers two advantages.
First, the recovery strategies contribute additively---explicit
extraction captures high-precision named references, contextual
discovery finds alias-based and functional references, anaphoric
resolution recovers pronoun-based references, and abbreviated matching
catches partial names---without interference or ordering constraints
between strategies.
Second, the separation of knowledge acquisition from link recovery
allows each knowledge source to be refined or replaced independently
without affecting the recovery strategies.

% ---------------------------------------------------------------------------
\begin{figure}[t]
\centering
\small
\begin{verbatim}
    Layer 1: Knowledge Acquisition (concurrent)
  ┌──────────┐  ┌──────────┐  ┌──────────┐
  │  Model    │  │ Document │  │ Explicit │
  │ Analysis  │  │ Analysis │  │Ref. Extr.│
  └─────┬────┘  └────┬─────┘  └────┬─────┘
        │             │             │
        ├─────────────┤             │
        ▼             │             │
  ┌──────────┐        │             │
  │ Pattern  │        │             │
  │ Learning │        │             │
  └─────┬────┘        │             │
        └──────┬──────┘             │
               │  shared knowledge  │
               ▼                    │
  ┌──────────────┐ ┌───────────┐   │
  │ Contextual   │ │ Anaphoric │   │
  │  Ref. Disc.  │ │ Ref. Res. │   │
  │+ Validation  │ │           │   │
  └──────┬───────┘ └─────┬─────┘   │
         │               │         │
         ├───────────────┤         │
         ▼               │         │
  ┌──────────┐           │         │
  │Abbrev.   │           │         │
  │Ref. Match│           │         │
  └────┬─────┘           │         │
       │  Layer 2: Link Recovery   │
       └──────────┬───────────────┘
                  ▼
        ┌──────────────────┐
        │ Merge + Boundary │
        │ Filter + Evidence│
        │     Filter       │
        └──────────────────┘
      Layer 3: Link Consolidation
\end{verbatim}
\caption{Fan-out / fan-in pipeline architecture. Layer~1 acquires knowledge
concurrently; pattern learning depends on model analysis and runs after the
initial analyses. Layer~2 runs four reference-type-specific recovery strategies
that converge at the merge point. Explicit reference extraction has no
knowledge dependencies and runs in Layer~1.}
\label{fig:pipeline}
\end{figure}
% ---------------------------------------------------------------------------

\subsection{Knowledge Acquisition}\label{sec:knowledge}

Layer~1 builds a shared knowledge base from the raw document and model.
Model analysis, document analysis, and explicit reference extraction
are mutually independent---each reads only the raw inputs---enabling
concurrent execution.
Pattern learning and multiword enrichment depend on model analysis
results and run once the initial analyses complete.

\subsubsection{Model Analysis}\label{sec:model-analysis}

We analyze component names from the architecture model to distinguish
\emph{architectural} names from \emph{ambiguous} ones.
Architectural names (e.g., \texttt{TaskScheduler}, \texttt{RenderEngine})
identify specific system roles, while ambiguous names
(e.g., \texttt{Core}, \texttt{Adapter}) are single common-English words
that documentation authors frequently use generically.

Classification uses a few-shot LLM prompt with examples drawn from
software-engineering textbook domains (compilers, operating systems,
networking), followed by a deterministic structural guard:
multi-word names, CamelCase compounds (detected via the regex pattern
\texttt{[a-z][A-Z]}), and all-uppercase acronyms are always classified
as architectural, regardless of LLM output.
This hybrid approach leverages LLM judgment for ambiguous cases while
guaranteeing that structurally unambiguous identifiers are never
misclassified.

The analysis additionally extracts: (a)~\emph{implementation relationships}
between component names (e.g., \texttt{BasicScheduler}~$\to$~\texttt{Scheduler}),
identified by shared substrings; (b)~\emph{shared vocabulary} across
components, mapping words that appear in multiple component names;
and (c)~\emph{generic partial words}, the set of ambiguous-classified
words that also appear as subwords of other component names, used later
to calibrate abbreviated reference matching.

\subsubsection{Document Analysis}\label{sec:doc-analysis}

Document analysis builds an \emph{alias table}---a mapping from
alternative names to their canonical component---that downstream phases
use to recognize references the raw component list alone cannot match.
The document text itself is never modified; only the mapping is produced.

\paragraph{Why alias discovery matters.}
Architecture documents rarely refer to components exclusively by their
canonical model names.
Authors abbreviate (\emph{AST} for \texttt{AbstractSyntaxTree}),
use trailing words as shorthand (\emph{Dispatcher} for
\texttt{EventDispatcher}), or introduce synonyms (\emph{task runner}
for \texttt{TaskScheduler}).
Without an alias table, any strategy that matches only canonical
names will systematically miss these references, capping recall.

\paragraph{Alternative name discovery.}
An LLM extracts three categories of aliases from the document:
\begin{itemize}
    \item \emph{Abbreviations}: short forms introduced via parenthetical
      definitions (e.g., ``Abstract Syntax Tree (AST)'' defines AST).
    \item \emph{Synonyms}: alternative names that unambiguously identify
      exactly one component.
    \item \emph{Partial references}: trailing words of multi-word component
      names used alone (e.g., ``Dispatcher'' for \texttt{EventDispatcher}).
\end{itemize}

\noindent A calibrated few-shot judge validates each proposed mapping.
The judge auto-approves abbreviations formed from component-name initials,
trailing words of multi-word names, and CamelCase identifiers, while
rejecting generic terms (\emph{system}, \emph{process}).
A deterministic \emph{CamelCase rescue} overrides any judge rejection of
CamelCase identifiers, since constructed identifiers are never generic
English words.

\paragraph{The challenge of partial references.}
Among the three alias types, partial references require the most
careful handling because they sit at the intersection of two
conflicting goals.
On the one hand, trailing-word shorthands are pervasive in
architecture documents---authors naturally write ``Server'' after
introducing ``HTML5 Server''---so ignoring them sacrifices recall.
On the other hand, many trailing words are ordinary English words
(\emph{server}, \emph{manager}, \emph{core}) that appear frequently
in non-referential contexts, creating false positives if matched
indiscriminately.
Abbreviations and synonyms do not share this tension:
abbreviations are structurally distinctive (uppercase acronyms),
and synonyms are validated for uniqueness.
Partial references, by contrast, are often single common words
whose referential status depends on context.

This ambiguity motivates a three-step pipeline that progressively
refines which partials are safe to use and in what context:

\begin{enumerate}
    \item \emph{Statistical enrichment.}
      For each multi-word component, if its trailing word appears
      in at least three sentences where the full name is absent,
      and no other component shares that trailing word,
      the partial mapping is added automatically.
      For trailing words classified as ambiguous by model analysis,
      the match requires capitalized occurrences to avoid
      coincidental overlap with generic uses.
      This step catches patterns the LLM missed by relying on
      distributional evidence rather than semantic judgment.
    \item \emph{Usage classification.}
      Per-partial LLM calls determine whether each single-word
      generic partial is typically used as a component reference
      (\textsc{name}) or as an ordinary English word
      (\textsc{ordinary}) in the document.
      For example, ``Server'' might be classified as \textsc{name}
      in a document that consistently capitalizes it when referring
      to the component, but as \textsc{ordinary} in a document
      where ``server'' appears generically dozens of times.
    \item \emph{Downstream gating.}
      \textsc{Ordinary}-classified partials remain in the alias table
      for deterministic link creation
      (abbreviated reference matching, \S\ref{sec:partial}),
      but are excluded from the confirmed-alias context provided
      to the Layer~3 boundary filter.
      This prevents the filter from treating generic uses as
      validated component references while still allowing the
      deterministic matcher to produce candidate links that the
      filter can then evaluate individually.
\end{enumerate}

\subsubsection{Pattern Learning}\label{sec:patterns}

Two independent LLM passes identify \emph{subprocess terms}---words that
refer to internal parts of components rather than the components themselves
(e.g., ``garbage collector'' inside a \texttt{MemoryManager}).
The first pass proposes candidates; the second cross-validates them in a
debate structure, rejecting terms that may be valid component references.
This debate mechanism reduces the risk of over-filtering in later
validation phases.
The phase also extracts \emph{action indicators} (verbs used when a
component performs an action) and \emph{effect indicators} (verbs
describing results), which serve as contextual signals during validation.

Pattern learning reads the filtered component list (excluding
implementation names identified by model analysis) and the document
sentences.
This dependency places pattern learning after model analysis,
forming a brief enrichment step between knowledge acquisition and
link recovery.

% ---------------------------------------------------------------------------

\subsection{Link Recovery}\label{sec:recovery}

Layer~2 comprises four complementary strategies, each targeting a
distinct reference type that the others cannot efficiently capture:

\begin{itemize}
    \item \textbf{Explicit reference extraction} recovers verbatim
      component names using only raw inputs (no Layer~1 knowledge),
      optimizing for precision at the cost of recall.
    \item \textbf{Contextual reference discovery} uses the full
      knowledge base (aliases, synonyms, partial names) to recover
      references that require contextual understanding---functional
      descriptions, interaction participants, and known aliases---then
      validates candidates through a multi-step pipeline.
    \item \textbf{Anaphoric reference resolution} resolves pronominal
      and definite-description references (``it'', ``the service'')
      back to previously mentioned components.
    \item \textbf{Abbreviated reference matching} deterministically
      matches abbreviated component names discovered during document
      analysis.
\end{itemize}

\noindent All four strategies' outputs converge only at the Layer~3
merge point.
Explicit extraction has no knowledge dependencies and runs
concurrently with Layer~1; the remaining three strategies consume the
shared knowledge base.

\subsubsection{Explicit Reference Extraction}\label{sec:seed}

Explicit extraction establishes a high-precision baseline before
any knowledge is available.
It accepts either an external baseline (e.g., TransArc CSV)
or runs a self-contained LLM-based extractor called \textbf{ILinker2}.
Because ILinker2 reads only the raw sentences and component
list---no abbreviations, synonyms, or partial references---it runs
concurrently with Layer~1 knowledge acquisition, overlapping its
LLM cost with the knowledge phases rather than waiting for them.

ILinker2 uses two complementary LLM passes over batched document segments
(50~sentences per batch with 5-sentence overlap), each with a
\emph{different prompt framing}:

\begin{itemize}
    \item \textbf{Pass~A (mention-framed):} ``For each sentence, identify
      which components are explicitly mentioned.'' This pass is optimized
      for exhaustive recall of named references.
    \item \textbf{Pass~B (actor-framed):} ``For each sentence, determine
      which component is the subject or primary actor.'' This reframing
      surfaces references that Pass~A may miss by focusing on agency
      rather than name matching.
\end{itemize}

\noindent Both passes share exclusion rules that reject names appearing
only inside dotted package paths (e.g., \texttt{renderer.utils.config}
does not reference \texttt{Renderer}) and generic English words used in
their ordinary sense.

The merge strategy applies \emph{asymmetric voting}:
exact-name matches from \emph{either} pass are accepted (union), while
non-exact matches (synonyms, partials) require agreement from \emph{both}
passes (intersection).
Exact matches are reliably identified by a single pass, while fuzzy
matches benefit from independent agreement to suppress hallucinations.

As a standalone extractor, ILinker2 achieves approximately 95\% precision
but only 76\% recall, deliberately trading recall for precision.
The recall gap---alias-based references, functional descriptions,
pronominal coreferences---is addressed by the three knowledge-enriched
strategies in Layer~2.

\subsubsection{Contextual Reference Discovery}\label{sec:entity}

Where explicit extraction finds only references recognizable from the
component name alone, contextual discovery leverages the full Layer~1
knowledge base to recover references that \emph{require} contextual
understanding.
The two strategies differ in three ways:

\begin{enumerate}
    \item \textbf{Knowledge-enriched prompts.}
      The discovery prompt receives all abbreviations, synonyms, and
      partial references from document analysis, enabling it to
      recognize aliases (e.g., ``AST'' for \texttt{AbstractSyntaxTree})
      and partial names (e.g., ``Dispatcher'' for
      \texttt{EventDispatcher}) that the knowledge-free explicit
      extractor cannot.
    \item \textbf{Broader inclusion scope.}
      Beyond direct name matches, the prompt covers six reference
      patterns: space-separated forms of compound names, functional
      descriptions, known aliases, interaction participants (``X sends
      data to Y'' references both X and Y), and passive/prepositional
      mentions (``handled by X'').
      Explicit extraction is limited to named and actor references.
    \item \textbf{Same-prompt intersection for variance reduction.}
      Rather than using two different prompt framings (as explicit
      extraction does), contextual discovery runs the \emph{same}
      prompt \emph{twice} independently and keeps only candidates
      found in \emph{both} passes (intersection).
      This eliminates spurious candidates that a single pass might
      over-extract due to LLM non-determinism---a risk that grows
      with the broader inclusion scope.
\end{enumerate}

\noindent An \emph{abbreviation guard} rejects invalid abbreviation
expansions by verifying that the context following the abbreviation is
consistent with the full component name.

For components that receive no links from either explicit extraction
or the discovery phase, a \emph{targeted recovery} step issues
single-component prompts with all known aliases, recovering references
missed by the broad-scope batch processing.

\paragraph{Validation.}
Candidates from contextual discovery undergo a four-step validation
process:

\begin{enumerate}
    \item \emph{Generic-mention detection:}
      For candidates where an ambiguous component name appears only in
      lowercase (e.g., ``core'' rather than ``Core''), an LLM pre-check
      determines whether the usage refers to the specific architectural
      component or is an ordinary English word.
      The LLM sees anchor sentences where the full name appears
      (for calibration) alongside the candidate sentences, and classifies
      each as \textsc{component} or \textsc{generic}.
      Generic-classified candidates are rejected before the main
      validation pipeline, preventing downstream false positives from
      coincidental word overlap.
    \item \emph{Code-first auto-approval:} Candidates where a component
      name or known alias appears verbatim in the sentence text (verified
      by word-boundary matching in code) are approved without LLM cost.
    \item \emph{Two-pass LLM intersection:} Remaining candidates are
      validated by two independent LLM passes with complementary
      foci---one on \emph{actor role} (``is the component performing
      an action?'') and one on \emph{direct reference} (``does the text
      refer to the specific architectural component?'').
      A candidate is accepted only if \emph{both} passes approve it,
      applying intersection voting for precision.
    \item \emph{Evidence post-filter:} For components classified as
      ambiguous by model analysis, an additional LLM pass must cite the
      \emph{exact substring} from the sentence that names the component.
      The cited text is verified against the actual sentence content in
      code, rejecting hallucinated evidence.
\end{enumerate}

\subsubsection{Anaphoric Reference Resolution}\label{sec:coref}

Sentences containing anaphoric expressions---pronouns (\emph{it},
\emph{its}, \emph{they}, \emph{their}, \emph{this}, \emph{these},
\emph{that}, \emph{those}) and definite descriptions (\emph{the
component}, \emph{the service})---are analyzed to resolve references
back to architecture components.

The pipeline uses a unified \emph{cases-in-context} strategy:
each pronoun-bearing sentence is presented as an individual case with
a $\pm5$-sentence bidirectional context window, processed in batches
of~10.
The LLM sees each case's surrounding context (marked with
\texttt{>>>} for the target sentence) and resolves any pronoun
that refers to an architecture component.
This per-case presentation with local context is robust across both
simple and complex documents, eliminating the need for document
complexity gating.

The resolution enforces a \emph{citation-grounded verification}
protocol: the LLM must cite an antecedent sentence where the component
was explicitly named, along with the exact quote containing the
component name.
This citation is verified deterministically in code:
\begin{enumerate}
    \item The component name (or a known alias) must appear verbatim in
      the cited antecedent sentence.
    \item The antecedent must be within 3~sentences of the pronoun.
    \item Sentences identified as subprocess descriptions by pattern
      learning are excluded.
\end{enumerate}

\noindent This hybrid LLM-plus-code verification eliminates a class of
hallucinated resolutions where the LLM fabricates a plausible but
nonexistent antecedent.
Because the verification is strict (verbatim name match within
3~sentences), false positives are effectively eliminated at the code
level, making anaphoric links safe to accept without further review in
Layer~3.

\subsubsection{Abbreviated Reference Matching}\label{sec:partial}

For each partial reference learned during document analysis
(e.g., ``Dispatcher''~$\to$~\texttt{EventDispatcher}), the pipeline
deterministically scans all sentences for word-boundary matches,
producing links for any sentence-component pairs not already covered
by the preceding strategies.
A \emph{clean mention} filter ensures the partial name appears as a
standalone word---not inside a dotted path (\texttt{x.Dispatcher.init}),
a hyphenated compound, or adjacent to a period indicating a qualified
name.

This strategy is entirely deterministic (no LLM calls), providing
stable recall for abbreviated references at zero additional inference
cost.
Its output is filtered by the convention-aware boundary filter in
Layer~3.

% ---------------------------------------------------------------------------

\subsection{Link Consolidation}\label{sec:merge}

Layer~3 merges the outputs of all recovery strategies and applies
two filtering stages to remove false positives.

\subsubsection{Priority-Based Merge}\label{sec:dedup}

Links from all four strategies are deduplicated by
$(sentence, component)$ pair.
When multiple strategies produce the same link, the highest-priority
source is retained according to a fixed ranking:
$\text{explicit} > \text{contextual} > \text{anaphoric} > \text{abbreviated}$.
A \emph{parent-overlap guard} removes links to implementation
components when the abstract parent is already linked to the same
sentence (e.g., if both \texttt{Scheduler} and
\texttt{BasicScheduler} are linked to a sentence, the implementation
variant is dropped).

\subsubsection{Convention-Aware Boundary Filter}\label{sec:boundary}

A structured LLM filter applies a three-step reasoning guide to
\emph{all} links, regardless of provenance:

\begin{enumerate}
    \item \textbf{Hierarchical name reference:}
      Reject if the component name appears only as a prefix in a
      dotted or qualified path (e.g., ``\texttt{X.utils} provides
      helper functions'' does not reference component~X).
      An exception preserves links when the sentence also names
      the component with the word ``component''
      (e.g., ``for the X component'').
    \item \textbf{Name confusion:}
      Reject if the name is used as a technology, methodology, or
      generic English word in a non-architectural sense.
      A critical exception preserves links when a component is
      \emph{named after} a technology (e.g., a component called
      ``Kafka Broker'')---any sentence describing that technology's
      capabilities is about the component.
    \item \textbf{Default:}
      If neither rejection pattern applies, the link is kept.
\end{enumerate}

\noindent Structural guardrails ensure multi-word component names and
CamelCase identifiers are never treated as generic words, and that
interaction sentences produce links for all participating components.

\subsubsection{Evidence Filter}\label{sec:evidence-filter}

After boundary filtering, a lightweight deterministic
\emph{evidence filter} removes links that lack textual grounding.
A link $s \to c$ is retained if \emph{any} of the following holds:

\begin{enumerate}
    \item The component name appears as a standalone word-boundary
      match in the sentence text (case-sensitive for single-word names,
      case-insensitive for multi-word names), excluding occurrences
      inside dotted paths or hyphenated compounds.
    \item A known alias (synonym, abbreviation, or partial reference
      from document analysis) appears in the sentence text.
    \item The link originates from explicit extraction, which already
      verified the named reference.
    \item The link originates from anaphoric resolution, which already
      verified the antecedent citation.
\end{enumerate}

\noindent Links that satisfy none of these conditions---where the
component name is absent from the sentence and no alias or prior
verification provides evidence---are dropped.
This filter requires zero LLM calls and removes residual false
positives from contextual discovery when the LLM hallucinated a
reference that does not appear in the text.

The rationale for accepting anaphoric links without additional review
is that antecedent verification (\S\ref{sec:coref}) already provides
strong evidence: the component was explicitly named in a nearby sentence
and the pronoun grammatically refers back to it.
Empirically, applying an LLM judge to anaphoric links removes true
positives but never catches false positives, as the code-level
verification is more reliable than LLM re-evaluation.

% ---------------------------------------------------------------------------

\subsection{Cross-Cutting Design Principles}\label{sec:principles}

Three principles guide the pipeline design across all layers:

\paragraph{Asymmetric voting.}
We apply \emph{intersection} voting (both passes must agree) for
precision-critical decisions: explicit extraction merge for non-exact
matches (\S\ref{sec:seed}) and contextual discovery validation
(\S\ref{sec:entity}).
This reflects the observation that false positives in early layers
propagate through the pipeline and are difficult to remove later.

\paragraph{Citation-grounded verification.}
Rather than trusting LLM judgments as opaque decisions, we require the
LLM to produce \emph{verifiable evidence}---antecedent sentence citations
in anaphoric resolution, exact-substring evidence in validation---that
is checked deterministically in code.
This hybrid approach leverages LLM reasoning for semantic understanding
while using programmatic guards to catch hallucinations.
The pattern applies in two places: anaphoric antecedent verification
(\S\ref{sec:coref}) and the evidence post-filter for ambiguous names
(\S\ref{sec:entity}).
A complementary but purely deterministic guard---the abbreviation guard
(\S\ref{sec:entity})---validates abbreviation expansions via string
matching without LLM involvement.

\paragraph{Benchmark-clean prompt design.}
All few-shot examples, reasoning guides, and prompt templates use
abstract placeholders (X, Y) or safe software-engineering textbook
domains (compilers, operating systems, networking).
No benchmark component names, dataset-specific vocabulary, or
gold-standard information appears in any prompt, ensuring that
measured performance reflects genuine generalization rather than
data leakage.
